---
title: "Statistical Rethinking Chapter 3 Practice"
author: "Dag Tanneberg"
date: "4/1/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("rethinking")
```


```{r}
p_grid <- seq(0, 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, 9, prob = p_grid)
posterior <- prior * likelihood
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
```

# Easy problems

## 3E1
```{r}
mean(samples < .2)
```

## 3E2
```{r}
mean(samples > .8)
```

## 3E3
```{r}
mean(samples > .2 & samples < .8)
```

## 3E4
```{r}
p_grid[max(which(cumsum(posterior) <= .2))]
quantile(samples, .2)
```

## 3E5
```{r}
p_grid[min(which(cumsum(posterior) >= .8))]
quantile(samples, .8)
```

## 3E6
```{r}
HPDI(samples, .66)
```

## 3E7
```{r}
PI(samples, .66)
alpha <- 1 - .66
quantile(samples, prob = c(alpha / 2, 1 - alpha / 2))
```

# Medium problems
## EM1
```{r}
likelihood <- dbinom(8, 15, p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

## EM2
```{r}
samples <- sample(p_grid, 10e3, prob = posterior, replace = TRUE)
HPDI(samples, prob = .9)
```

## EM3
```{r}
dummy_data <- rbinom(10e3, 15, samples)
mean(dummy_data == 8)
```

## EM4
```{r}
dummy_data <- rbinom(10e3, 9, samples)
mean(dummy_data == 6)
```

# Hard Problems

```{r}
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,
    0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,
    1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,
    1,0,1,1,1,0,1,1,1,1
)
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,
    1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,
    1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,
    0,0,0,1,1,1,0,0,0,0
)
```

## 3H1.

Using grid approximation, compute the posterior distribution for the
probability of a birth being a boy. Assume a uniform prior probability. Which
parameter value maximizes the posterior probability?

```{r}
grid_size <- 1000
p_grid <- seq(0, 1, length.out = grid_size)
prior <- dunif(.5, 0, 1)
likelihood <- dbinom(x = sum(birth1, birth2), size = length(c(birth1, birth2)),
    prob = p_grid
)
posterior <- prior * likelihood / sum(likelihood)
p_grid[which.max(posterior)]
plot(p_grid, posterior, type = "l")
```

## 3H2.

Using the sample function, draw 10,000 random parameter values from the
posterior distribution you calculated above. Use these samples to estimate the
50%, 89%, and 97% highest posterior density intervals.

```{r}
p_samples <- sample(p_grid, 10e3, prob = posterior, replace = TRUE)
sapply(c(.50, .89, .97), function(l){HPDI(p_samples, prob = l)})
```

## 3H3.

Use rbinom to simulate 10,000 replicates of 200 births. You should end up with
10,000 numbers, each one a count of boys out of 200 births. Compare the
distribution of predicted numbers of boys to the actual count in the data (111
boys out of 200 births). There are many good ways to visualize the simulations,
but the dens command (part of the rethinking package) is probably the easiest
way in this case. Does it look like the model fits the data well? That is, does
the distribution of predictions include the actual observation as a central,
likely outcome?

```{r}
replicated_births <- rbinom(10e4, size = 200, prob = p_samples)
dens(replicated_births); abline(v = sum(birth1, birth2))
```

The distribution of replicated male births centers over the empirically observed
count. The distribution symmetrically falls off to either side. Consequently,
this distribution seems to fit the data well.

## 3H4.

Now compare 10,000 counts of boys from 100 simulated first borns only to the
number of boys in the first births, birth1. How does the model look in this
light?

```{r}
replicated_first_births <- rbinom(10e4, size = 100, prob = p_samples)
plot(density(replicated_first_births)); abline(v = sum(birth1))
```

The distribution noticeably overestimates the probability of a firstborn son.

## 3H5.

The model assumes that sex of first and second births are independent.
To check this assumption, focus now on second births that followed female first
borns. Compare 10,000 simulated counts of boys to only those second births
that followed girls. To do this correctly, you need to count the number of first
borns who were girls and simulate that many births, 10,000 times. Compare the
counts of boys in your simulations to the actual observed count of boys
following girls. How does the model look in this light? Any guesses what is
going on in these data?

```{r}
replicated_2nd_births <- rbinom(10e4, size = length(birth1) - sum(birth1), prob = p_samples)
plot(density(replicated_2nd_births)); abline(v = sum(birth2[!birth1]))
```

Parents who got a girl first are a lot more likely to have a boy next. Whereas
the distribution of first borns is much more balanced between genders, female
first borns seem to trigger some sex selection process.
