---
title: Learning Tasks
author: Dag Tanneberg
date: 04/14/2019
output: html_document
---

## Introduction

- mlr works on tasks
- tasks encapsulate data & information about the ml problem
- there is a virtual superclass Task()
- Existing tasks:
    + RegrTask()
    + ClassifTask()
    + SurvTask()
    + ClusterTask()
    + MultilabelTask()
    + CostSensTask(): cost-sensitive classification

## Regression Task

```{r}
library("mlr")
data(BostonHousing, package = "mlbench")
regr.task <- makeRegrTask(id = "bh", data = BostonHousing, target = "medv")
regr.task
```

## Classification Task

- target must be a factor

```{r}
data(BreastCancer, package = "mlbench")
df <- BreastCancer; df$Id <- NULL
classif.task <- makeClassifTask(id = "bc", data = df, target = "Class")
classif.task
```

## Survival Analysis

- declaration expects time & censoring variable
- argument <censoring> defaults to "rcens" (right censoring)

```{r}
data(lung, package = "survival")
str(lung)
lung$status <- (lung$status == 2)
surv.task <- makeSurvTask(data = lung, target = c("time", "status"))
surv.task
```

## Multilabel classification

- each object can belong to more than one category
- data must have as many logical target columns as there are labels
- column names are taken as class label

```{r}
yeast <- getTaskData(yeast.task) # extract data from a task object
labels <- colnames(yeast)[seq(14)]
yeast.task <- makeMultilabelTask(id = "multilabel", data = yeast, target = labels)
yeast.task
```

## Cluster Analysis

- unsupervised learning $\rightarrow$ no target required

```{r}
data(mtcars, package = "datasets")
cluster.task <- makeClusterTask(data = mtcars)
cluster.task
```

## Cost-Sensitive Classification

- assumes differential costs for classification errors
- in Cost-Sensitive Classification costs depend on the *specific training example*
- each example $(x, y)$ has a cost vector $K$ with length $K$ = # of classes

```{r}
df <- iris
cost <- matrix(runif(150 * 3, 0, 2000), 150) * (1 - diag(3))[df$Species,]
df$Species = NULL

costsens.task = makeCostSensTask(data = df, cost = cost)
costsens.task
```

